---
title: "Running External Isoform Predictors (CPAT, SignalP, Pfam) on Windows and Linux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running External Isoform Predictors on Windows and Linux}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`run_isoform_switch(..., run_predictors = TRUE)` can enrich an
`IsoformSwitchAnalyzeR` result with three external predictions that are not
pure-R tools:

| Tool                    | Predicts                          | R import function       |
|--------------------------|-----------------------------------|--------------------------|
| CPAT (`cpat`/`run_cpat.py`) | Coding potential                | `analyzeCPAT()`          |
| SignalP (`signalp`/`signalp6`) | Signal peptides              | `analyzeSignalIP()`     |
| InterProScan / hmmscan  | Pfam protein domains              | `analyzeInterProScan()` / `analyzePFAM()` |

These tools only ship Linux binaries. ExpressOM handles this transparently on
every platform:

* **Native Linux / macOS** (including an R session already running *inside*
  WSL): the tools are called directly through the local `bash` -- no extra
  configuration needed beyond having them on `PATH` (or in a conda
  environment named `isoform_tools`).
* **Windows**: the same commands are transparently routed through
  `wsl -d <distro> bash ...`, so you keep the familiar R/RStudio-on-Windows
  workflow while the heavy lifting happens inside a WSL Ubuntu distribution.

You do not need to write different code for each platform -- `use_wsl`
defaults to `TRUE` on Windows and is ignored everywhere else.

## 1. Checking your environment first

Before running a full analysis, use `debug_wsl()` to verify that the tools,
conda environment, and reference databases are reachable. This works
identically on Windows and Linux/macOS:

```{r}
library(ExpressOM)

# On Windows (routes through WSL by default):
debug_wsl(distro = "Ubuntu-22.04", out_dir = "./results")

# On native Linux / macOS (no distro needed):
debug_wsl(use_wsl = FALSE, out_dir = "./results")
```

This prints a checklist and writes a machine-readable summary to
`results/Log/wsl_debug.json`:

```
Checking external predictor tools in the native execution environment...
  ✓ execution environment reachable
  ✓ conda found
  ✓ conda environment 'isoform_tools' exists
  ✓ cpat
  ✗ run_cpat.py
  ✓ signalp6
  ✗ signalp
  ✓ hmmscan
  ✗ interproscan.sh
  ✓ Pfam-A.hmm
  ✓ CPAT logit models
```

If `wsl_available` is `FALSE`, the underlying shell itself could not be
reached (missing `wsl` executable on Windows, or no `bash` on PATH on
Linux/macOS) -- fix that first. If `conda_env_exists` is `FALSE`, the tools
are still tried directly on `PATH`, but a warning is issued so you know why a
tool might be missing.

## 2. Installing the tools

### Windows (via WSL)

```{r}
install_wsl_isoform_tools(
  distro              = "Ubuntu-22.04",
  use_mamba           = TRUE,
  install_databases   = TRUE,
  windows_signalp_dir = "C:/tools/signalp-6.0"  # optional, if you have a licensed copy
)
```

### Linux / macOS

Install the tools with conda/mamba directly (no WSL layer involved):

```bash
mamba create -n isoform_tools -c bioconda cpat hmmer signalp
conda activate isoform_tools
```

Then fetch the CPAT/Pfam reference data with the same helper used on
Windows:

```{r}
install_isoform_databases(use_wsl = FALSE)
```

## 3. Running the predictors

```{r}
switch_list <- run_isoform_switch(
  isoform_obj    = isoform_import,
  condition      = "condition",
  level          = "Treated",
  base           = "Control",
  fasta_file     = "reference/transcripts.fa",
  gff_file       = "reference/annotation.gtf",
  out_dir        = "results/isoform",
  run_predictors = TRUE,

  # Windows only -- ignored on Linux/macOS:
  use_wsl        = TRUE,
  wsl_distro     = "Ubuntu-22.04",

  # Optional: cap CPU threads used by hmmscan / InterProScan.
  # Leave NULL to auto-detect (all cores minus one).
  predictor_cpu  = 4
)
```

Internally, `run_isoform_switch()`:

1. Calls `debug_wsl()` once up front and stops with a clear error if the
   execution shell itself (WSL, or native `bash`) cannot be reached, or
   warns (without stopping) if the `isoform_tools` conda environment is
   missing but tools might still be found directly on `PATH`.
2. Extracts NT/AA FASTA sequences from the `SwitchList`.
3. Runs CPAT on the NT sequences, then SignalP and Pfam (InterProScan first,
   falling back to `hmmscan`) on the AA sequences.
4. Imports each tool's results back into the `SwitchList` via the matching
   `IsoformSwitchAnalyzeR::analyze*()` function.
5. Checkpoints the result to `step3_predictors.rds` inside `save_dir`, so a
   crashed or interrupted run can resume without repeating already-finished
   predictor steps.

## 4. Performance notes

* `predictor_cpu` controls the `--cpu` / `-cpu` flag passed to `hmmscan` and
  `interproscan.sh`. Leaving it `NULL` auto-detects available cores via
  `parallel::detectCores()` and reserves one core for the R session itself.
* Conda-environment discovery (`conda.sh` lookup) is cached for the duration
  of the R session, so calling `run_isoform_switch()` repeatedly (e.g. once
  per contrast in a loop, as described in the main package README for
  subgroup analyses) does not repeat an expensive filesystem-wide `find` each
  time.
* CPAT, SignalP, and Pfam are still run sequentially per call, since each
  step is itself already multi-threaded; if you are looping over many
  contrasts, consider parallelising *across* contrasts with
  `BiocParallel::bplapply()` instead, giving each worker its own `out_dir`.

## 5. Troubleshooting

| Symptom | Likely cause |
|---|---|
| `debug_wsl()` reports `wsl_available = FALSE` on Windows | WSL is not installed, or the named `distro` does not exist. Run `wsl --list` in a terminal to check. |
| `debug_wsl()` reports `wsl_available = FALSE` on Linux/macOS | No `bash` on `PATH`. This should not normally happen; check your shell configuration. |
| A specific tool shows `✗` even though it is installed | It may only exist inside a conda environment with a different name than `isoform_tools`, or your shell's login profile does not source `conda.sh` automatically. Pass the correct `conda_env` name, or activate the environment before launching R. |
| `Pfam-A.hmm not found` | Run `install_isoform_databases()` (with `use_wsl = TRUE` on Windows or `FALSE` on Linux/macOS). |
| Predictor step silently skipped | Every predictor step is optional and fails soft: if a tool or database is missing, that section is skipped with a `message()` rather than stopping the whole pipeline. Check the console output and `results/isoform/Log/wsl_commands.log` for details. |
