---
title: "DTE/DTU Analysis Report (`r params$level` vs `r params$base`)"
author: "ExpressOM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
    highlight: tango
params:
  level:             "treated"
  base:               "control"
  top_n:              15
  genes_of_interest:  !r NULL
  dte_csv:            ""
  dtu_csv:            ""
  dexseq_csv:         !r NULL
  has_dexseq:         FALSE
  has_switch:         FALSE
  plot_dir:           "plots"
  switch_plot_top_n:  10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(DT)
library(ggplot2)

# Emit a markdown image for `basename` if it exists under params$plot_dir,
# preferring a .png rendition (browsers can't display a PDF through an <img>
# tag) and falling back to the .pdf if no PNG conversion is available.
# Called from `results='asis'` chunks; returns invisibly if the file is
# missing so callers can just call it unconditionally.
.fig <- function(basename, cap = NULL) {
  png_path <- file.path(params$plot_dir, sub("\\.pdf$", ".png", basename))
  pdf_path <- file.path(params$plot_dir, basename)
  rel <- if (file.exists(png_path)) file.path("plots", sub("\\.pdf$", ".png", basename))
         else if (file.exists(pdf_path)) file.path("plots", basename)
         else return(invisible(NULL))
  cat(sprintf("\n![%s](%s)\n\n", if (is.null(cap)) "" else cap, rel))
  invisible(TRUE)
}

# List files (basenames) directly under a subdirectory of params$plot_dir,
# or character(0) if it doesn't exist.
.list_plot_files <- function(subdir, pattern = "\\.(pdf|png)$") {
  d <- file.path(params$plot_dir, subdir)
  if (!dir.exists(d)) return(character(0))
  list.files(d, pattern = pattern, recursive = TRUE)
}
```

## Overview

- **Comparison**: `r params$level` vs `r params$base`
- **Date**: `r Sys.time()`
- **Top N shown**: `r params$top_n`
- **Genes of interest**: `r if (length(params$genes_of_interest) > 0) paste(params$genes_of_interest, collapse = ", ") else "None requested"`

## DTE Results (Differential Transcript Expression)

### Volcano Plot
```{r dte-volcano, results='asis'}
.fig("DTE_volcano.pdf", "Volcano plot")
```

### MA Plot
```{r dte-ma, results='asis'}
.fig("DTE_MA.pdf", "MA plot")
```

### Top Significant Transcripts
```{r dte-topbar, results='asis'}
.fig("DTE_top_barplot.pdf", "Top DE transcripts")
```

### Interactive Table (top 1000 by padj)
```{r dt-dte}
dte_data <- read.csv(params$dte_csv, check.names = FALSE)
dte_data <- dte_data[order(dte_data$padj), ][seq_len(min(1000, nrow(dte_data))), ]
DT::datatable(dte_data, filter = "top", options = list(scrollX = TRUE))
```

## DTU Results (Differential Transcript Usage)

### P-value Distribution
```{r dtu-hist, results='asis'}
.fig("DTU_pvalue_hist.pdf", "DTU p-value histogram")
```

### Interactive Table
```{r dt-dtu}
dtu_data <- read.csv(params$dtu_csv, check.names = FALSE)
DT::datatable(dtu_data, filter = "top", options = list(scrollX = TRUE))
```

```{r dexseq-section, echo=FALSE, results='asis', eval=isTRUE(params$has_dexseq)}
cat("\n## DEXSeq Results (Differential Transcript/Exon Usage)\n\n")
cat("DEXSeq treats each transcript as an \"exonic bin\" within its parent gene and ",
    "tests for differential usage independently of the DRIMSeq-based DTU test above; ",
    "the two are commonly compared as a cross-validation of DTU calls.\n\n", sep = "")
cat("### Interactive Table\n\n")
```

```{r dt-dexseq, eval=isTRUE(params$has_dexseq)}
dexseq_data <- read.csv(params$dexseq_csv, check.names = FALSE)
dexseq_data <- dexseq_data[order(dexseq_data$padj), ]
DT::datatable(dexseq_data, filter = "top", options = list(scrollX = TRUE))
```

```{r switch-overview-header, echo=FALSE, results='asis', eval=isTRUE(params$has_switch)}
cat("\n## Isoform Switch Overview\n\n")
cat("### dIF vs. Isoform Switch Significance\n\n")
```

```{r switch-overview-fig, results='asis', eval=isTRUE(params$has_switch)}
.fig("IsoformSwitch_overview.pdf", "Isoform Switch Overview")
```

```{r switch-genomewide, echo=FALSE, results='asis', eval=isTRUE(params$has_switch)}
# Genome-wide functional-consequence and alternative-splicing summaries
# (extractConsequenceSummary / extractConsequenceEnrichment /
# extractSplicingSummary / extractSplicingEnrichment). Each is optional --
# only shown if the corresponding file exists, since some can be legitimately
# skipped for small/uniform datasets (see safe_run() calls in R/mod_isoform.R).
genome_wide_figs <- list(
  list(file = "ConsequenceSummary.pdf",    title = "Functional Consequences (Summary)",
       cap  = "Genome-wide summary of functional consequences of isoform switching"),
  list(file = "ConsequenceEnrichment.pdf", title = "Functional Consequences (Enrichment)",
       cap  = "Enrichment of gain vs. loss for each functional consequence"),
  list(file = "SplicingSummary.pdf",       title = "Alternative Splicing (Summary)",
       cap  = "Genome-wide summary of alternative splicing event types"),
  list(file = "SplicingEnrichment.pdf",    title = "Alternative Splicing (Enrichment)",
       cap  = "Enrichment of gain vs. loss for each alternative splicing event type")
)
present <- vapply(genome_wide_figs, function(f) file.exists(file.path(params$plot_dir, f$file)) ||
                    file.exists(file.path(params$plot_dir, sub("\\.pdf$", ".png", f$file))),
                  logical(1))
if (any(present)) {
  cat("### Genome-wide Functional Consequences & Alternative Splicing\n\n")
  for (f in genome_wide_figs[present]) {
    cat("#### ", f$title, "\n\n", sep = "")
    .fig(f$file, f$cap)
  }
}
```

```{r switch-top, echo=FALSE, results='asis', eval=isTRUE(params$has_switch)}
top_switch_files <- .list_plot_files("top_switches", pattern = "\\.(pdf|png)$")
# Once PDFs are converted to PNG we'd otherwise list both extensions for the
# same figure; keep one entry per figure, preferring the PNG.
top_switch_files <- unique(sub("\\.pdf$", ".png", top_switch_files))
if (length(top_switch_files) > 0) {
  cat("### Top ", params$switch_plot_top_n, " Isoform Switches\n\n", sep = "")
  for (fn in top_switch_files) {
    fig_label <- tools::file_path_sans_ext(basename(fn))
    cat("#### ", fig_label, "\n\n", sep = "")
    cat(sprintf("\n![](%s)\n\n", file.path("plots", "top_switches", fn)))
  }
}
```

## Gene-specific Plots

```{r gene-specific, echo=FALSE, results='asis'}
genes <- params$genes_of_interest
if (length(genes) == 0) {
  cat("\nNo custom genes requested.\n")
} else {
  gene_plot_types <- list(
    list(prefix = "proportions_", label = "**Transcript proportions**"),
    list(prefix = "DEXSeq_",      label = "**DEXSeq transcript usage**",
         only_if = isTRUE(params$has_dexseq)),
    list(prefix = "SwitchPlot_",  label = "**Isoform switch summary (structure, expression, usage)**"),
    list(prefix = "Sashimi_",     label = "**Sashimi-style junction usage**"),
    list(prefix = "ExonUsage_",   label = "**Exon-bin usage comparison**")
  )
  for (g in genes) {
    section_lines <- character(0)
    for (pt in gene_plot_types) {
      if (!is.null(pt$only_if) && !isTRUE(pt$only_if)) next
      fn <- paste0(pt$prefix, g, ".pdf")
      png_exists <- file.exists(file.path(params$plot_dir, sub("\\.pdf$", ".png", fn)))
      pdf_exists <- file.exists(file.path(params$plot_dir, fn))
      if (png_exists || pdf_exists) {
        section_lines <- c(section_lines, pt$label, "")
      }
    }
    if (length(section_lines) > 0) {
      cat("\n### ", g, "\n\n", sep = "")
      for (pt in gene_plot_types) {
        if (!is.null(pt$only_if) && !isTRUE(pt$only_if)) next
        fn <- paste0(pt$prefix, g, ".pdf")
        if (file.exists(file.path(params$plot_dir, sub("\\.pdf$", ".png", fn))) ||
            file.exists(file.path(params$plot_dir, fn))) {
          cat(pt$label, "\n\n", sep = "")
          .fig(fn)
        }
      }
    }
  }
}
```
