---
title: "DTE/DTU Analysis Report (`r params$level` vs `r params$base`)"
author: "ExpressOM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
    highlight: tango
params:
  level: "treated"
  base: "control"
  top_n: 15
  genes_of_interest: !r NULL
  dte_csv: ""
  dtu_csv: ""
  dexseq_csv: !r NULL
  has_dexseq: FALSE
  has_switch: FALSE
  plot_dir: "plots"
  switch_plot_top_n: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(DT)
})

`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0) y else x
}

.md_path <- function(x) {
  gsub(" ", "%20", x, fixed = TRUE)
}

.read_csv_safe <- function(path) {
  if (is.null(path) || length(path) == 0 || !nzchar(path)) {
    return(NULL)
  }

  if (!file.exists(path)) {
    return(NULL)
  }

  tryCatch(
    utils::read.csv(path, check.names = FALSE),
    error = function(e) NULL
  )
}

.plot_abs <- function(fname, subdir = NULL) {
  if (is.null(subdir)) {
    file.path(params$plot_dir, fname)
  } else {
    file.path(params$plot_dir, subdir, fname)
  }
}

.plot_exists <- function(fname, subdir = NULL) {
  png_name <- sub("\\.pdf$", ".png", fname)

  file.exists(.plot_abs(png_name, subdir)) ||
    file.exists(.plot_abs(fname, subdir))
}

.fig <- function(fname, cap = NULL, subdir = NULL) {
  png_name <- sub("\\.pdf$", ".png", fname)

  png_abs <- normalizePath(.plot_abs(png_name, subdir), winslash = "/", mustWork = FALSE)
  pdf_abs <- normalizePath(.plot_abs(fname, subdir), winslash = "/", mustWork = FALSE)

  caption <- if (is.null(cap) || !nzchar(cap)) {
    tools::file_path_sans_ext(basename(fname))
  } else {
    cap
  }

  # NOTE: this Rmd is knit from a temp copy outside the report's own
  # directory, and html_document defaults to self_contained = TRUE, so
  # pandoc embeds each image at *knit* time relative to that temp
  # location -- a "plots/..." relative link (the old behavior here) never
  # resolves and the image is silently dropped, even though the file
  # exists on disk once the report is moved into place. Using the
  # already-computed absolute path sidesteps that entirely and matches
  # the pattern already used successfully in
  # dge_regionreport_customcode.Rmd's .expressom_fig().
  if (file.exists(png_abs)) {
    cat(sprintf("\n![%s](%s)\n\n", caption, .md_path(png_abs)))
    return(invisible(TRUE))
  }

  if (file.exists(pdf_abs)) {
    cat(sprintf("\n[Download PDF: %s](%s)\n\n", caption, .md_path(pdf_abs)))
    return(invisible(TRUE))
  }

  cat(sprintf("\n*Figure not available: %s*\n\n", fname))
  invisible(FALSE)
}

.list_switch_figs <- function(subdir) {
  d <- file.path(params$plot_dir, subdir)

  if (!dir.exists(d)) {
    return(character(0))
  }

  files <- list.files(
    d,
    pattern = "\\.(pdf|png)$",
    recursive = TRUE
  )

  if (length(files) == 0) {
    return(character(0))
  }

  bases <- unique(tools::file_path_sans_ext(files))

  out <- vapply(
    bases,
    function(b) {
      png <- paste0(b, ".png")
      pdf <- paste0(b, ".pdf")

      if (file.exists(file.path(d, png))) {
        png
      } else if (file.exists(file.path(d, pdf))) {
        pdf
      } else {
        NA_character_
      }
    },
    character(1),
    USE.NAMES = FALSE
  )

  out[!is.na(out)]
}

.list_top_switch_figs <- function() .list_switch_figs("top_switches")
.list_by_consequence_figs <- function() .list_switch_figs("by_consequence")
```

# Overview

- **Comparison:** `r params$level` vs `r params$base`
- **Date:** `r Sys.time()`
- **Top N shown:** `r params$top_n`
- **Genes of interest:** `r if (length(params$genes_of_interest) > 0) paste(params$genes_of_interest, collapse = ", ") else "None requested"`

```{r skipped_plots_section, results='asis'}
skip_path <- file.path(dirname(params$plot_dir), "skipped_plots.csv")
skip_df <- .read_csv_safe(skip_path)

if (!is.null(skip_df) && nrow(skip_df) > 0) {
  cat("\n## Skipped / removed plots\n\n")

  cat(
    "The following plots were skipped or removed because they were empty, corrupt, or had no underlying data.\n\n"
  )

  cat(knitr::kable(skip_df, format = "markdown"), sep = "\n")
  cat("\n")
}
```

# DTE Results (Differential Transcript Expression)

## Volcano Plot

```{r dte_volcano, results='asis'}
.fig("DTE_volcano.pdf", "DTE volcano plot")
```

## MA Plot

```{r dte_ma, results='asis'}
.fig("DTE_MA.pdf", "DTE MA plot")
```

## Top Significant Transcripts

```{r dte_top, results='asis'}
.fig("DTE_top_barplot.pdf", "Top differentially expressed transcripts")
```

## Interactive Table (top 1000 by adjusted p-value)

```{r dte_table}
dte_data <- .read_csv_safe(params$dte_csv)

if (is.null(dte_data)) {
  knitr::asis_output("*DTE result file not found.*")
} else {
  if ("padj" %in% colnames(dte_data)) {
    dte_data <- dte_data[order(dte_data$padj, na.last = TRUE), ]
  }

  dte_data <- utils::head(dte_data, 1000)

  DT::datatable(
    dte_data,
    filter = "top",
    options = list(
      scrollX = TRUE,
      pageLength = 10
    )
  )
}
```

# DTU Results (Differential Transcript Usage)

## P-value Distribution

```{r dtu_hist, results='asis'}
.fig("DTU_pvalue_hist.pdf", "DTU p-value histogram")
```

## Top Significant Transcripts

```{r dtu_top, results='asis'}
.fig("DTU_top_barplot.pdf", "Top DTU transcripts")
```

## Interactive Table

```{r dtu_table}
dtu_data <- .read_csv_safe(params$dtu_csv)

if (is.null(dtu_data)) {
  knitr::asis_output("*DTU result file not found.*")
} else {
  if ("adj_pvalue" %in% colnames(dtu_data)) {
    dtu_data <- dtu_data[order(dtu_data$adj_pvalue, na.last = TRUE), ]
  } else if ("pvalue" %in% colnames(dtu_data)) {
    dtu_data <- dtu_data[order(dtu_data$pvalue, na.last = TRUE), ]
  }

  DT::datatable(
    dtu_data,
    filter = "top",
    options = list(
      scrollX = TRUE,
      pageLength = 10
    )
  )
}
```

## Cross-engine Concordance

```{r engine_concordance, results='asis'}
if (.plot_exists("Engine_concordance.pdf")) {
  cat("\nSignificant gene concordance across DTE/DTU/DEXSeq engines:\n\n")
  .fig("Engine_concordance.pdf", "Cross-engine concordance")
}
```

```{r dexseq_header, results='asis', eval=isTRUE(params$has_dexseq)}
cat("\n# DEXSeq Results (Differential Transcript/Exon Usage)\n\n")

cat(
  "DEXSeq treats each transcript as an exonic bin within its parent gene and ",
  "tests for differential usage independently of the DRIMSeq-based DTU test above. ",
  "The two are commonly compared as cross-validation of DTU calls.\n\n",
  sep = ""
)

cat("## Interactive Table\n\n")
```

```{r dexseq_table, eval=isTRUE(params$has_dexseq)}
dexseq_data <- .read_csv_safe(params$dexseq_csv)

if (is.null(dexseq_data)) {
  knitr::asis_output("*DEXSeq result file not found.*")
} else {
  order_col <- intersect(
    c("gene_qvalue", "padj", "pvalue"),
    colnames(dexseq_data)
  )[1]

  if (!is.na(order_col)) {
    dexseq_data <- dexseq_data[order(dexseq_data[[order_col]], na.last = TRUE), ]
  }

  DT::datatable(
    dexseq_data,
    filter = "top",
    options = list(
      scrollX = TRUE,
      pageLength = 10
    )
  )
}
```

```{r switch_header, results='asis', eval=isTRUE(params$has_switch)}
cat("\n# Isoform Switch Overview\n\n")
cat("## dIF vs. Isoform Switch Significance\n\n")
```

```{r switch_overview, results='asis', eval=isTRUE(params$has_switch)}
.fig("IsoformSwitch_overview.pdf", "Isoform switch overview")
```

```{r genome_wide_switch_plots, results='asis', eval=isTRUE(params$has_switch)}
genome_wide_figs <- list(
  list(
    file = "ConsequenceSummary.pdf",
    title = "Functional Consequences (Summary)",
    cap = "Genome-wide summary of functional consequences of isoform switching"
  ),
  list(
    file = "ConsequenceEnrichment.pdf",
    title = "Functional Consequences (Enrichment)",
    cap = "Enrichment of gain vs. loss for each functional consequence"
  ),
  list(
    file = "SplicingSummary.pdf",
    title = "Alternative Splicing (Summary)",
    cap = "Genome-wide summary of alternative splicing event types"
  ),
  list(
    file = "SplicingEnrichment.pdf",
    title = "Alternative Splicing (Enrichment)",
    cap = "Enrichment of gain vs. loss for each alternative splicing event type"
  )
)

present <- vapply(
  genome_wide_figs,
  function(f) .plot_exists(f$file),
  logical(1)
)

if (any(present)) {
  cat("\n## Genome-wide Functional Consequences & Alternative Splicing\n\n")

  for (f in genome_wide_figs[present]) {
    cat("### ", f$title, "\n\n", sep = "")
    .fig(f$file, f$cap)
  }
}
```

```{r top_switches_section, results='asis', eval=isTRUE(params$has_switch)}
top_switch_files <- .list_top_switch_figs()

if (length(top_switch_files) > 0) {
  cat("\n## Top ", params$switch_plot_top_n, " Isoform Switches\n\n", sep = "")

  for (fn in top_switch_files) {
    fig_label <- tools::file_path_sans_ext(basename(fn))

    cat("### ", fig_label, "\n\n", sep = "")
    .fig(fn, fig_label, subdir = "top_switches")
  }
}
```

```{r by_consequence_section, results='asis', eval=isTRUE(params$has_switch)}
by_consequence_files <- .list_by_consequence_figs()

if (length(by_consequence_files) > 0) {
  cat(
    "\n## Isoform Switches by Functional Consequence\n\n",
    "The same switch-plot type as above, but filtered to switches with at least one ",
    "identified functional consequence (domain/signal-peptide/coding-potential/NMD change) ",
    "and grouped by which consequence it is.\n\n",
    sep = ""
  )

  for (fn in by_consequence_files) {
    fig_label <- tools::file_path_sans_ext(basename(fn))

    cat("### ", fig_label, "\n\n", sep = "")
    .fig(fn, fig_label, subdir = "by_consequence")
  }
}
```

# Gene-specific Plots

```{r gene_plots, results='asis'}
genes <- params$genes_of_interest

if (length(genes) == 0) {
  cat("\nNo custom genes requested.\n")
} else {
  gene_plot_types <- list(
    list(
      prefix = "proportions_",
      label = "**Transcript proportions**"
    ),
    list(
      prefix = "DEXSeq_",
      label = "**DEXSeq transcript usage**",
      only_if = isTRUE(params$has_dexseq)
    ),
    list(
      prefix = "SwitchPlot_",
      label = "**Isoform switch summary (structure, expression, usage)**",
      only_if = isTRUE(params$has_switch)
    ),
    list(
      prefix = "Sashimi_",
      label = "**Sashimi-style junction usage**",
      only_if = isTRUE(params$has_switch)
    ),
    list(
      prefix = "ExonUsage_",
      label = "**Exon-bin usage comparison**",
      only_if = isTRUE(params$has_switch)
    )
  )

  for (g in genes) {
    section_lines <- character(0)

    for (pt in gene_plot_types) {
      if (!is.null(pt$only_if) && !isTRUE(pt$only_if)) next

      fn <- paste0(pt$prefix, g, ".pdf")

      if (.plot_exists(fn)) {
        section_lines <- c(section_lines, pt$label)
      }
    }

    if (length(section_lines) > 0) {
      cat("\n## ", g, "\n\n", sep = "")

      for (pt in gene_plot_types) {
        if (!is.null(pt$only_if) && !isTRUE(pt$only_if)) next

        fn <- paste0(pt$prefix, g, ".pdf")

        if (.plot_exists(fn)) {
          plain_label <- gsub("\\*\\*", "", pt$label)

          cat(pt$label, "\n\n", sep = "")
          .fig(fn, paste(g, "-", plain_label))
        }
      }
    }
  }
}
```